---
title: "Analysis of electric vehicle usage patterns in New Zealand"
subtitle: "Summary Statistical Report"
author: "Rafferty Parker and Ben Anderson (University of Otago)"
date: 'Last run at: `r Sys.time()`'
output:
  bookdown::html_document2:
    code_folding: hide
    fig_caption: yes
    number_sections: yes
    self_contained: no
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::pdf_document2:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
  bookdown::word_document2:
    fig_caption: yes
    toc: yes
    toc_depth: 2
bibliography: EVBBmendeleyrefs.bib  # Won't knit with the bibliography for some reason. Seems to be a problem with pandoc as opposed to the .bib file.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(readr) # tidy data reading
library(data.table) # cos we like data.table (you may not in which case dplyr is fine :-)
library(lubridate) # for data & time manip
library(hms) # for hh:mm:ss if we need it
library(ggplot2) # fancy plots
library(ggjoy)
library(dplyr) # for filter
library(forcats)
library(knitr) # for knitting
library(kableExtra) # for extra kable

# colour blind palettes for charts
# http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#a-colorblind-friendly-palette
# with grey
cbgPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
# with black
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# data file to use
file <- "EVBB_processed_all_v1.0_20180125.csv"
# for Mac
user <- Sys.info()[[7]]
if(user == "ben"){
  dPath <- "/Volumes/hum-csafe/Research Projects/GREEN Grid/externalData/flipTheFleet/safe/testData/2019_01_25/"
  dFile <- paste0(dPath, file, ".zip") # use zipped data
  if(!file.exists(dFile)) {
    # we probably don't have the HCS mounted so switch to local
    dPath <- "~/Data/NZ_GREENGrid/ftf/"
    dFile <- paste0(dPath, file, ".zip")
  }
} else {
  # for Xubuntu:
  dPath <- "/run/user/1001/gvfs/smb-share:server=storage.hcs-p01.otago.ac.nz,share=hum-csafe,user=student%5Cparra358/Research Projects/GREEN Grid/externalData/flipTheFleet/safe/testData/2019_01_25/"
  dFile <- paste0(dPath, "EVBB_processed_all_v1.0_20180125.csv")
}

print(paste0("Using ", dFile))
rawDF <- readr::read_csv(dFile) # creates a tidyverse tibble https://www.tidyverse.org/articles/2018/01/tibble-1-4-1/

```


```{r dataPrep1}
df <- rawDF # so can always re-create df without having to re-load data
# don't do anything else here to avoid confusion
```

```{r dataPrep2}
#Combine date and time columns into POSIXct datetime
df$dateTime <- lubridate::as_datetime(paste0(df$date, df$time))

# set correct order for days of the week
df$day_of_week <- ordered(df$day_of_week, levels=c("Monday", "Tuesday", "Wednesday",
                                                   "Thursday", "Friday", "Saturday", "Sunday"))
# set charge type
df$chargeType <- ifelse(df$charge_power_kw > 0, "Standard charge", NA) 
df$chargeType <- ifelse(df$charge_power_kw >= 7, "Fast charge", df$chargeType)
df$chargeType <- ifelse(is.na(df$chargeType), "Not charging", df$chargeType) # not charging

# set charge type order so charts make sense from left (std) to right (fast)
df$chargeType <- cut(df$charge_power_kw, c(-Inf, 0.01, 7, Inf), labels = c('Not charging', 'Standard charging', 'Fast charging'))
df$chargeType <- factor(df$chargeType, ordered = TRUE)

# Rename vehicle ids to something more user-friendly
df$dvID <- factor(df$id, ordered = TRUE)
levSeq <- seq(1:length(levels(df$dvID)))
levSeqChar <- as.character(levSeq)
df$dvID <- factor(df$dvID,
  labels = levSeqChar)
df$dvID <- as.character(df$dvID)
df$dvID <- paste("Vehicle", df$dvID, sep = " ")

names(df)[names(df) == 'state_of_charge_percent'] <- 'SoC_percent'

df$qHour <- hms::trunc_hms(df$time, 15*60) # truncate to previous 15 min

# Month as ordered factor
df$month <- factor(df$month, ordered = TRUE, levels = c("Jan", "Feb", "Mar", "Apr", "May",
                                                        "Jun", "Jul", "Aug", "Sep", "Oct",
                                                        "Nov", "Dec"))


# Create factor for weekdays/weekends
weekdays1 <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
df$weekday <- factor((df$day_of_week %in% weekdays1), 
                   levels = c(TRUE, FALSE), labels = c('Weekday', 'Weekend'), ordered = TRUE)

# removal of silly state of charge percentage values
df$SoC_percent[df$SoC_percent > 100] <- NA
df$SoC_percent[df$SoC_percent < 0] <- NA

# removal of silly charge_power_kw values
# "...charging stations are being developed with capacities of 120kW in New Zealand"
# (Concept Consulting report)
df$charge_power_kw[df$charge_power_kw > 120] <- NA
```

```{r dataTable stuff, results="hide"}
dt <- as.data.table(df) # creates a data.table for fast data crunching

# Remove vehicles with all-zero charging values
summaryDT <- dt[, .(mean = mean(charge_power_kw), sd = sd(charge_power_kw), nObs = .N), keyby = .(dvID)]
includeDT <- summaryDT[mean != 0]
setkey(includeDT, dvID)
setkey(dt, dvID)
finalDT <- dt[includeDT]
dt <- finalDT

#Create charge flag
dt <- dt[, chargeFlag := ifelse(shift(charge_power_kw) == 0 & charge_power_kw > 0 & shift(charge_power_kw, type = "lead") > 0,
                                "first", "Not charging"), by = id] 
# the previous value was 0 but this value and the next value > 0

dt <- dt[, chargeFlag := ifelse(shift(charge_power_kw) > 0 & charge_power_kw > 0, 
                                "charging", chargeFlag), by = id] 
# previous value > 0, this value > 0

dt <- dt[, chargeFlag := ifelse(shift(charge_power_kw == 0, type = "lead") & charge_power_kw > 0 & shift(charge_power_kw) > 0, 
                                "last", chargeFlag), by = id] 
# next value = 0, this value and the previous value > 0



dt$chargeFlag <- ordered(dt$chargeFlag, levels=c("first", "charging", "last"))
 table(dt$chargeFlag, useNA = "always")

# ATTN BEN - delete following lines if not necessary
#dt <- dt[ , `:=`( chargeCount = .N ) , by = chargeFlag ]

#dt <- dt[, obsDiffTime := difftime(dateTime,shift(dateTime)), by = id] # time since previous observation (within id)

#dt <- dt[, obsDiffSecs := as.numeric(obsDiffTime)] # seconds since previous observation (within id) - could include reset to 0 after midnight
  
chargingDT <- dt[charge_power_kw > 0] # select just charging


```


```{r establishment of chargeBegins and chargeEnds dataframes}
chargeBegins <- chargingDT[chargingDT$chargeFlag == "first" , ]
chargeEnds <- chargingDT[chargingDT$chargeFlag == "last", ]
```

```{r calculateChargeDurations}
# select the observations which we've flagged as first & last in a sequence of charging
firstLastDT <- dt[chargeFlag == "first" | chargeFlag == "last"]
# flag the first of a pair
firstLastDT <- firstLastDT[, pairOK := ifelse(chargeFlag == "first" & shift(chargeFlag == "last", type = "lead"), "Pair start", NA)]
# flag the second of a pair
firstLastDT <- firstLastDT[, pairOK := ifelse(chargeFlag == "last" & shift(chargeFlag == "first"), "Pair end", pairOK)]
# calculate the time diff between all obs
firstLastDT <- firstLastDT[, pairDuration := difftime(time1 = dateTime, time2 = shift(dateTime), units = c("mins"))]
# we only want the time difference which was calculated for an obs where pairOK == "Pair end". This should also be where chargeFlag == "last" _except_ for where we have no 'first' (e.g. at start of data)
# note that we will still have pairs that bridge 00:00 which will give us -ve values
# if we have a -ve value then we need to change the calculation to add the time
# up to midnight from the start to the time after midnight to the end
firstLastDT <- firstLastDT[pairOK == "Pair start" & shift(pairDuration < 0, type = "lead"), 
                           toMidnight := difftime(time1 = as.hms("23:59:59"), time2 = time)]
firstLastDT <- firstLastDT[pairOK == "Pair end" & pairDuration < 0, 
                           afterMidnight := difftime(time1 = time, time2 = as.hms("00:00:00"), units = c("mins"))]
firstLastDT <- firstLastDT[, pairDurationFix := shift(toMidnight) + afterMidnight]
firstLastDT <- firstLastDT[, pairDurationFinal := ifelse(pairDuration <0,
                                                         pairDurationFix,
                                                         pairDuration)]


# Remove overly large values
firstLastDT <- firstLastDT[pairDurationFinal < 6000]

# ATTN BEN do we do the following here or disply the plots below that depend on the small values and then remove them for further analysis?

# Remove standard charges of duration less than 8 mins
firstLastDT <- firstLastDT[!(firstLastDT$pairDurationFinal < 8 & firstLastDT$chargeType == "Standard charging"),]

# Remove (lone) fast charge of duration greater than 10 hours
firstLastDT <- firstLastDT[!(firstLastDT$pairDurationFinal > 840 & firstLastDT$chargeType == "Fast charging"),]

# ATTN BEN rather than specify ...[pairOK == "Pair end", ] for all the following firstLastDT$pairDurationFinal plots, should we just remove all rows that are not "Pair end"?

```


# Note

Based on and inspired by the [UK DoT statistical report 2018](https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/764270/electric-chargepoint-analysis-2017-domestics.pdf).

This is a summary report. A discussion into the background around the effects of EV charging in New Zealand, as well as detailed information on data limitations and data cleaning procedure is given in the full report.

# Data information{#data}

## Background

The data consisted of `r nrow(dt)` data points (of which `r nrow(chargingDT)` were charging (power demand > 0)) from `r  length(levels(as.factor(df$dvID)))` vehicles over 8 months (April 2018 - January 2019) derived from FlipTheFleet's [blackbox recorder](https://flipthefleet.org/ev-black-box/). The recorder provided measurements at 1 minute frequency of charging behaviour and battery charge state.

Due to privacy considerations, the data is not publically available.

## Definitions:

The capacity of most domestic charging is between 1.8kW to 7kW, whereas charging power above 7kW is available at purpose-built charging stations[@ConceptConsulting2018]. Each charging event was therefore seperated into "Fast" (> = 7kW) and "Standard" (below 7kW). 

A charging event was defined as a continuous sequence of 1 minute observations per vehicle when > 0 kW of demand was observed.

## Cleaning and Preparation
_Perhaps for the summary report this entire subsection could be removed?_
There were `r nrow(summaryDT) - nrow(includeDT)` vehicles within the provided data that had no recorded charging occur. These were immediately discarded.

Some instances of charging power greater than 120kW were recorded. These were considered anomolies and discarded, as these exceed the capacity of the highest charging stations available in New Zealand [@ConceptConsulting2018].

Instances of battery state of charge being greater than 100% or less than 0% were also discarded.

Standard charge durations of less than 8 minutes were frequently encountered near the end of a longer charging cycle, where the state of charge had reached it's maximum. These were assumed to be minor 'top-ups', and were discarded. In addition, slow charging events greater than 100 hours were discarded, as were fast charge durations greater than 14 hours. These were presumed anomolies as they exceed the battery capacity of electric vehicles commonly encountered in New Zealand.

For more detailed information on the data cleaning process refer to the main report.


# Key Findings:

```{r keyFindings}
stdMedian <- median(chargingDT[chargeType == "Standard charging"]$charge_power_kw, na.rm = TRUE)
stdMean <- mean(chargingDT[chargeType == "Standard charging"]$charge_power_kw, na.rm = TRUE)
  
fastMedian <- median(chargingDT[chargeType == "Fast charging"]$charge_power_kw, na.rm = TRUE)
fastMean <- mean(chargingDT[chargeType == "Fast charging"]$charge_power_kw, na.rm = TRUE)
```

 * _Power supplied_: The median power supplied during a standard charging was `r round(stdMedian,2)`  kW. The mean was slightly higher at `r round(stdMean,2)` kW. Fast charging observations had a higher median of `r round(fastMedian,2)` kW (mean = `r round(fastMean,2)`);
  * _Charging duration_: Charging durations tended to fall into one of two groups - longer 'overnight' charges with a median of XX hours and shorter events during the day both at standard and fast charge rates with a median duration of XX hours. 
  * _Time of Day_: charging events were more frequent at specific times of the day and day of the week with more evening and over-night charging during weekdays and more day-time charging at weekends. The power demand also varied according to time of day and day of the week.

# Observed demand

Figure \@ref(fig:obsPower) shows the distribution of observed charging kW demand by inferred charge type. This plot shows that fast charges are relatively rare in the dataset whilst standard charges are much more common, and are concentrated around 1.8kW, 3kW and 6kW. 

```{r obsPower, fig.cap="Observed power demand distribution by charge type"}
p <- ggplot2::ggplot(chargingDT, aes(x = charge_power_kw, fill = chargeType, binwidth = 0.1)) +
  geom_histogram() +
  facet_wrap(. ~ chargeType, scales = "free")

# now draw the plot with any fancy extras we want
p + labs(y = "Density",
       x = "Power (kW)") +
  guides(fill = guide_legend(title = "Charge type:")) +
  scale_fill_manual(values=cbgPalette) + # use colour-blind friendly palette
  theme(legend.position = "bottom")
```

```{r stdChargeSizing}
stdQT <- quantile(chargingDT[chargeType == "Standard charging"]$charge_power_kw)
fastQT <- quantile(chargingDT[chargeType == "Fast charging"]$charge_power_kw)
```

75% of standard charging observations were `r round(stdQT[[2]],2)` kW or more but the figure was `r round(fastQT[[2]],2)` kW or more for fast charging

# Daily demand

```{r dailyPower, fig.cap="Observed power demand distribution by day of the week and charge type"}
p <- ggplot2::ggplot(filter(dt, chargeType == "Standard charging" | chargeType == "Fast charging"), aes(x = day_of_week, colour = chargeType, group = day_of_week)) +
  geom_boxplot(aes(y = charge_power_kw)) +
  facet_wrap(. ~ chargeType, scales= "free_y")

p + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  labs(x = "Day of week",
       y = "Power (kW)") +
  guides(colour = guide_legend(title = "Charge type:")) +
  scale_colour_manual(values=cbgPalette) + # use colour-blind friendly palette
  theme(legend.position = "bottom")
```

Figure \@ref(fig:dailyPower) shows the distribution of observed charging kW demand by day of the week. We can see that fast charging varies in demand but standard charging is relatively constant across days.


# Charging duration {#duration}

Figure \@ref(fig:durationHist) shows the overall distribution of charging sequences. As would be expected, fast charging events tend to have a much shorted duration than standard charging.

```{r durationDescTableReduced}
t <- firstLastDT[pairOK == "Pair end" & pairDurationFinal > 8, 
                 .(N = .N,
                   mean = mean(pairDurationFinal),
                   median = median(pairDurationFinal),
                   min = min(pairDurationFinal),
                   max = max(pairDurationFinal)), 
                 keyby = .(chargeType)]
kableExtra::kable(t, 
                  caption = "Duration (minutes) of charge sequences by charge type", digits = 2) %>%
  kable_styling()
```


```{r durationHist, fig.cap="Duration of charging sequences"}

ggplot2::ggplot(firstLastDT[pairOK == "Pair end"], 
                aes(x = pairDurationFinal)) +
  geom_histogram(binwidth = 5) +
  facet_wrap(chargeType ~ ., scales = "free") +
  labs(x = "Minutes")
```



```{r durationTimeBox, fig.cap="Duration by time of charging start for sequences > 8 minutes"}

ggplot2::ggplot(firstLastDT[pairOK == "Pair end" & pairDurationFinal > 8 & pairDurationFinal < 6000], 
                aes(x = qHour, y = pairDurationFinal, group = qHour)) +
  geom_boxplot() +
  facet_grid(chargeType ~ ., scales = "free") +
  labs(x = "Time of Day",
       y = "Minutes")
```

```{r durationTimeMean, fig.cap="Mean duration (within quarter hours) by time of charging start for sequences > 8 minutes"}

plotDT <- firstLastDT[pairOK == "Pair end",
                      .(meanDuration = mean(pairDurationFinal)),
                      keyby = .(qHour, chargeType)]
ggplot2::ggplot(plotDT, 
                aes(x = qHour, y = meanDuration, colour = chargeType)) +
  geom_point() +
  labs(x = "Time of Day",
       y = "Minutes")
```


```{r meanDurationTable}
kableExtra::kable(t, caption = "Mean duration of charge events by charge type")
```

```{r meanDurationPlot, fig.cap="Mean duration by time of charging start"}
plotDT <- firstLastDT[pairOK == "Pair end", .(meanDuration = mean(pairDurationFinal, na.rm = TRUE)), keyby = .(chargeType, dateTime)]
```


# Time of charging

```{r chargeTime, fig.cap="Count of observed charging events by type, day of week and time"}
plotDT <- chargingDT[, .(count = .N), keyby = .(qHour, chargeType, day_of_week)]

# make a weekend facet label
plotDT <- plotDT[, weekEnd := "Weekend"]
plotDT <- plotDT[day_of_week != "Saturday" & day_of_week != "Sunday", weekEnd := "Week day"]

p <- ggplot2::ggplot(plotDT, aes(x = qHour, y = count, colour = day_of_week)) +
  geom_line() +
  facet_grid(weekEnd ~  chargeType, scales = "free_y")
  
p + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  labs(x = "Time of day",
       y = "Count") +
  guides(colour = guide_legend(title = "Day of week:")) +
  scale_colour_manual(values=cbgPalette) + # use colour-blind friendly palette
  theme(legend.position = "bottom")
```

Figure \@ref(fig:chargeTime) shows the distribution of observed charging by time of day and day of the week. Aggregating counts in this way emphasises the times at which charging most commonly occurs and we can see...
 
Fig: profile of median charging demand by time of day and day of the week  faceted by at home vs not at home

Charging demand varies somewhat by time of day and day of the week. Weekdays show ... whilst weekends show. Saturdays and Sundays vary with... 

```{r boxplotCharging, fig.cap="Boxplot of daily standard charging demand"}
p <- ggplot2::ggplot(subset(df, chargeType %in% "Standard charging"), 
                     aes(x = qHour, group = qHour, y = charge_power_kw)) +
  scale_colour_manual(values=cbbPalette) + # use colour-blind friendly palette
  geom_boxplot() # <- make the plot in an object first

p + labs(x = "Time of Day", y = "Power (kW)")
```


```{r plot3, fig.cap="Boxplot of daily fast charging demand"}
p <- ggplot2::ggplot(subset(dt, chargeType %in% "Fast charging"), 
                     aes(x = qHour, group = qHour, y = charge_power_kw)) +
  scale_colour_manual(values=cbbPalette) + # use colour-blind friendly palette
  geom_boxplot() # <- make the plot in an object first

p + labs(x = "Time of Day", y = "Power (kW)")
```



```{r chargeBeginsWeekday, fig.cap="Density plot of charging start times during weekdays"}
p <- ggplot(chargeBegins[chargeBegins$weekday == "Weekday", ], aes(x = qHour, fill = chargeType)) +
  geom_density(alpha = 0.3) 
  facet_grid(~weekday)
p + labs(x = "Time", fill = "Charge type")

```

```{r chargeBeginsWeekend, fig.cap="Density plot of charging start times during weekends"}
p <- ggplot(chargeBegins[chargeBegins$weekday == "Weekend", ], aes(x = qHour, fill = chargeType)) +
  geom_density(alpha = 0.3) 
p + labs(x = "Time", fill = "Charge type")

```

```{r chargeEndsWeekday, fig.cap="Density plot of charging end times during weekdays"}
p <- ggplot(chargeEnds[chargeEnds$weekday == "Weekday", ], aes(x = qHour, fill = chargeType)) +
  geom_density(alpha = 0.3) 
p + labs(x = "Time", fill = "Charge type" )

```


```{r chargeEndsWeekend, fig.cap="Density plot of charging end times during weekends"}
p <- ggplot(chargeEnds[chargeEnds$weekday == "Weekend", ], aes(x = qHour, fill = chargeType))  +
  scale_colour_manual(values=cbbPalette)+
  geom_density(alpha = 0.3) +
  facet_grid(~weekday)
p + labs(x = "Time", fill = "Charge type")

```



Standard charging events tended to begin at HH:MM during weekdays and HH:MM at weekends. 

Standard charging has a noticeably different profile to charging patterns for fast charges. It suggests that it is common for plug-in vehicle owners to charge overnight at home, and perhaps use the more powerful public chargepoints to top up during the day.

> Discuss any other patterns

# State of charge

The duration of charging events (see Section \@ref(duration)) suggests that EVs may be 'plugged in' at home (and elsewhere) for considerable durations. 


```{r SoCplot2, fig.cap= "Value of state of charge at beginning of charge"}
tmp <- chargeBegins[chargeBegins$SoC_percent < 90, ]
p <- ggplot(data=tmp, aes(tmp$SoC)) + geom_histogram(bins = 20)
p + labs(x = "State of charge when charging begins (%)")
ggsave("plots/SOC_when_charging_begins.png")
tmp <- NULL
```



Figure \@ref(fig:SoCplot2) shows that many vehicles arrive home with greater than 50% charge remaining and would therefore be able to transfer energy to the home during the evening grid peak as a form of demand response. 


Fig: Mean state of battery charge at the first 'at home' charging observation by hour and day of the week _No "at home" data with SOC_

> should show the timing of 'coming home' battery state?

Fig: Distribution of duration of charge events starting 'at home' in the evening (by day of the week) _Duration difficult to accurately determine without date due to charging occurring through the night_

The figure shows that vehicles may then be available for further demand response and/or re-charging for up to XX hours from this point.

> Discuss any other patterns

# References

