---
title: "EVBB Analysis"
author: "Rafferty Parker (University of Otago)"
date: 'Last run at: `r Sys.time()`'
output:
  bookdown::html_document2:
    code_folding: hide
    fig_caption: yes
    number_sections: yes
    self_contained: no
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    fig_caption: yes
    toc: yes
    toc_depth: 2
  bookdown::pdf_document2:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
subtitle: subtitle goes here
bibliography: refs.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(readr)
library(knitr)
library(lubridate)
library(hms)

iFile <- "data/EVBB_processed.csv" # set this here
```

# Introduction

When/where is the greatest amount of charging occurring? What is the power demand per vehicle/household/place of charging? What might this look like if 50% of households had an EV? What about if it matched current ICE ownership (eventually)?
  
Develop tools for analyzing the data so that data can be fed in and plots automatically constructed

# Brief literature review
What do we already know (or suspect) about when they will be charged. Include:

 * academic literature
 * consultant reports
 * etc

This is an example of a bibtex reference to a paper: [@stephenson_smart_2017]

This is an example of a reference to an R package: [@dplyr]

# Research questions

Break down the main questions & refer back to the literature

# Methods
## Data
Describe it:

 * where did you get it?
 * how many vehicles, where are they and what kinds?
 * anything we need to know about the data?
 * what did you have to do to it (cleaning etc)
 * descriptive analysis, e.g.:
   * Plot all charging with fast and slow charges differentiated
   * Plot fast charges with weekend and weekday differentiated 
   * Histograms of power (kW)

Load the data - note the feedback readr gives you on the assumed format of the columns. You might _not_ want to remove the original dateTime :-)

```{r load data}
df <- readr::read_csv(iFile)
```

This is a cross-references to Table \@ref(tab:tab1).

```{r tab1}

t <- summary(df)

knitr::kable(t, caption = "Data summary") # <- makes a pretty table
```


This is a cross-reference to Figure \@ref(fig:plot1).

```{r plot1, fig.cap="Density plot of charging power by car"}
p <- ggplot2::ggplot(df, aes(x = charge_power_kw)) +
  geom_density() # <- make the plot in an object first

p + labs(x = "Power (kW)") + facet_grid(id ~ .) # <- draw the plot and add more features
```


## Analysis

Analysis was conducted using R (`r R.version.string`) and the following packages:

 * ggplot2 [@ggplot2]
 * dplyr [@dplyr]

Reports were developed using knitr [@knitr] within bookdown [@bookdown].

# Charging Analysis

## Research question 1

When does charging happen?

```{r codeHalfHours}

df$halfHour <- hms::trunc_hms(df$time, 30*60) # <- code to half hours

```

This is a cross-reference to Figure \@ref(fig:plot2). Time is coded to half hours. 

```{r plot2, fig.cap="Boxplot of charging timing by car"}
p <- ggplot2::ggplot(df, aes(x = halfHour, group = halfHour, y = charge_power_kw)) +
  geom_boxplot() # <- make the plot in an object first

p + labs(x = "Time of Day", y = "Power (kW)") + facet_grid(day_of_week ~ id) # <- draw the plot and add more features
```

## Research question 2

# Conclusions

# References

