---
title: "Analysis of electric vehicle usage patterns in New Zealand"
subtitle: "Statistical Report"
author: "Rafferty Parker and Ben Anderson (University of Otago)"
date: 'Last run at: `r Sys.time()`'
output:
  bookdown::pdf_document2:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
  bookdown::html_document2:
    code_folding: hide
    fig_caption: yes
    number_sections: yes
    self_contained: no
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    fig_caption: yes
    toc: yes
    toc_depth: 2
#bibliography: EVBBmendeleyrefs.bib  # Won't knit with the bibliography for some reason. Seems to be a problem with pandoc as opposed to the .bib file.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(readr) # tidy data reading
library(data.table) # cos we like data.table (you may not in which case dplyr is fine :-)
library(lubridate) # for data & time manip
library(hms) # for hh:mm:ss if we need it
library(ggplot2) # fancy plots
library(ggjoy)
library(dplyr) # for filter
library(forcats) # used to reverse days of week in joy plots
library(knitr) # for knitting
library(kableExtra) # for extra kable

# colour blind palettes for charts
# http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#a-colorblind-friendly-palette
# with grey
cbgPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
# with black
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# data file to use
file <- "EVBB_processed_all_v1.0_20180125.csv"
# for Mac
user <- Sys.info()[[7]]
if(user == "ben"){
  dPath <- "/Volumes/hum-csafe/Research Projects/GREEN Grid/externalData/flipTheFleet/safe/testData/2019_01_25/"
  dFile <- paste0(dPath, file, ".zip") # use zipped data
  if(!file.exists(dFile)) {
    # we probably don't have the HCS mounted so switch to local
    dPath <- "~/Data/NZ_GREENGrid/ftf/"
    dFile <- paste0(dPath, file, ".zip")
  }
} else {
  # for Xubuntu:
  dPath <- "/run/user/1001/gvfs/smb-share:server=storage.hcs-p01.otago.ac.nz,share=hum-csafe,user=student%5Cparra358/Research Projects/GREEN Grid/externalData/flipTheFleet/safe/testData/2019_01_25/"
  dFile <- paste0(dPath, "EVBB_processed_all_v1.0_20180125.csv")
}

print(paste0("Using ", dFile))
rawDF <- readr::read_csv(dFile) # creates a tidyverse tibble https://www.tidyverse.org/articles/2018/01/tibble-1-4-1/

```


```{r dataPrep1}
df <- rawDF # so can always re-create df without having to re-load data
# don't do anything else here to avoid confusion
```

```{r dataPrep2}
#Combine date and time columns into POSIXct datetime
df$dateTime <- lubridate::as_datetime(paste0(df$date, df$time))

# set correct order for days of the week
df$day_of_week <- ordered(df$day_of_week, levels=c("Monday", "Tuesday", "Wednesday",
                                                   "Thursday", "Friday", "Saturday", "Sunday"))
# set charge type
df$chargeType <- ifelse(df$charge_power_kw > 0, "Standard charge", NA) 
df$chargeType <- ifelse(df$charge_power_kw >= 7, "Fast charge", df$chargeType)
df$chargeType <- ifelse(is.na(df$chargeType), "Not charging", df$chargeType) # not charging

# set charge type order so charts make sense from left (std) to right (fast)
df$chargeType <- cut(df$charge_power_kw, c(-Inf, 0.01, 7, Inf), labels = c('Not charging', 'Standard charging', 'Fast charging'))
df$chargeType <- factor(df$chargeType, ordered = TRUE)

# Rename vehicle ids to something more user-friendly
df$dvID <- factor(df$id, ordered = TRUE)
levSeq <- seq(1:length(levels(df$dvID)))
levSeqChar <- as.character(levSeq)
df$dvID <- factor(df$dvID,
  labels = levSeqChar)
df$dvID <- as.character(df$dvID)
df$dvID <- paste("Vehicle", df$dvID, sep = " ")

names(df)[names(df) == 'state_of_charge_percent'] <- 'SoC_percent'

df$qHour <- hms::trunc_hms(df$time, 15*60) # truncate to previous 15 min
#df$qHour <- format(as.POSIXct(hms::trunc_hms(df$time, 15*60)), "%H:%M")

# Month as ordered factor
df$month <- factor(df$month, ordered = TRUE, levels = c("Jan", "Feb", "Mar", "Apr", "May",
                                                        "Jun", "Jul", "Aug", "Sep", "Oct",
                                                        "Nov", "Dec"))


# Create factor for weekdays/weekends
weekdays1 <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
df$weekday <- factor((df$day_of_week %in% weekdays1), 
                   levels = c(TRUE, FALSE), labels = c('Weekday', 'Weekend'), ordered = TRUE)

# removal of silly state of charge percentage values
df$SoC_percent[df$SoC_percent > 100] <- NA
df$SoC_percent[df$SoC_percent < 0] <- NA

# removal of silly charge_power_kw values
# "...charging stations are being developed with capacities of 120kW in New Zealand"
# (Concept Consulting report)
df$charge_power_kw[df$charge_power_kw > 120] <- NA
```

```{r dataTable stuff, results="hide"}
dt <- as.data.table(df) # creates a data.table for fast data crunching

# Remove vehicles with all-zero charging values
summaryDT <- dt[, .(mean = mean(charge_power_kw), sd = sd(charge_power_kw), nObs = .N), keyby = .(dvID)]
includeDT <- summaryDT[mean != 0]
setkey(includeDT, dvID)
setkey(dt, dvID)
finalDT <- dt[includeDT]
dt <- finalDT

# Create charge flag
dt <- dt[, chargeFlag := ifelse(shift(charge_power_kw) == 0 & charge_power_kw > 0 & shift(charge_power_kw, type = "lead") > 0,
                                "first", "Not charging"), by = id] 
# the previous value was 0 but this value and the next value > 0

dt <- dt[, chargeFlag := ifelse(shift(charge_power_kw) > 0 & charge_power_kw > 0, 
                                "charging", chargeFlag), by = id] 
# previous value > 0, this value > 0

dt <- dt[, chargeFlag := ifelse(shift(charge_power_kw == 0, type = "lead") & charge_power_kw > 0 & shift(charge_power_kw) > 0, 
                                "last", chargeFlag), by = id] 
# next value = 0, this value and the previous value > 0



dt$chargeFlag <- ordered(dt$chargeFlag, levels=c("first", "charging", "last"))
 table(dt$chargeFlag, useNA = "always")

# ATTN BEN - delete following lines if not necessary
#dt <- dt[ , `:=`( chargeCount = .N ) , by = chargeFlag ]

#dt <- dt[, obsDiffTime := difftime(dateTime,shift(dateTime)), by = id] # time since previous observation (within id)

#dt <- dt[, obsDiffSecs := as.numeric(obsDiffTime)] # seconds since previous observation (within id) - could include reset to 0 after midnight
  
chargingDT <- dt[charge_power_kw > 0] # select just charging
```

# Introduction

The New Zealand government has set a target of increasing the number of EVs in New Zealand to 64,000 by 2021. High penetration of EVs would cause EV recharging to contribute a substantial portion of total electricity load. A report prepared for lines companies Orion, Powerco and Unison by Concept Consulting Group entitled "Driving change - Issues and options to maximise the opportunities from large-scale electric vehicle uptake in New Zealand" predicts that if all current light private vehicles were electric, annual residential electricity consumption would increase by approximately 30%, whereas if all vehicles including trucks were electric, this would increase the total electricity consumption of New Zealand by approximately 41%[concept_2018]. 

New Zealand's total electricity demand varies throughout the day, with weekdays in particular having two distinct "peaks"; one in the morning, and one in the evening. Providing the electicity to meet these demand peaks is a costly and inefficient process. Concurrent electric vehicle (EV) charging, especially in the early evening when many motorists return home, would have the potential to negatively impact the operation of the grid through drastically increasing peak loads [Azadfar2015], leading to an increased cost of electricity due to the requirement of expensive upgrades to the electricity grid[@stephenson_smart_2017].

This report hopes to provide further insight into the potential effects on the New Zealand electricity grid that may occur with a dramatic increase in EVs, so that these may be planned for and mitigated. It is based on and inspired by the [UK DoT statistical report 2018](https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/764270/electric-chargepoint-analysis-2017-domestics.pdf).


# Data information{#data}

## Background 

The data used has been provided by "Flip the Fleet", a community organisation that hopes to increase uptake of electric vehicles in New Zealand. Flip the Fleet have been collecting data on electric vehicle usage patterns, collected using Exact IOT Limited's [blackbox recorder](https://flipthefleet.org/ev-black-box/), a small electronic  device that connects to the vehicle's internal computer and sends detailed data about the battery health, consumption, speed, etc.

The data used consisted of `r nrow(dt)` data points from `r  length(levels(as.factor(dt$dvID)))` vehicles over 8 months (April 2018 - January 2019). The recorder provided measurements at 1 minute frequency of charging behaviour and battery charge state.

Due to privacy considerations, the data is not publically available.

## Initial cleaning

There were `r nrow(summaryDT) - nrow(includeDT)` vehicles in the data provided that had no recorded charging occur. These were immediately discarded.

Some instances of charging power greater than 120kW were recorded. These were considered anomolies and discarded, as these exceed the capacity of the highest charging stations available in New Zealand[@concept2018].

Instances of battery state of charge being greater than 100% or less than 0% were also discarded.

## Definitions and preparation

Charging data has been broadly seperated into two seperate catagories, "standard" and "fast". Standard charging is when the charger is reading less than 7kW - this is considered the upper limit of what can be obtained from a standard home charging scenario without an expensive wiring upgrade[@concept2018]. Fast charging is all charging above 7kW, and would likely occur at designated and purpose-built fast charging stations.

The data was also catagorised according to whether it was a weekday or not. This allows analysis to occur of differing charging patterns between weekdays and weekends, allowing for further accuracy in determining the effects of electric vehicles on grid peaks.

In order to determine charging durations, rows were initially flagged as "charging begins" if the charging power was greater than zero and the previous and following row's charging power were (respectively) equal to zero and greater than zero. Similarly, rows were flagged as "charge ends" if the charging power was greater than zero and the previous and following row's charging power were (respectively) greater than zero and equal to zero.


```{r establishment of chargeBegins and chargeEnds dataframes}
chargeBegins <- chargingDT[chargingDT$chargeFlag == "first" , ]
chargeEnds <- chargingDT[chargingDT$chargeFlag == "last", ]
```

Using this method we obtained `r nrow(chargeBegins)` instances of charge beginning, and `r nrow(chargeEnds)` instances of charge ending. The additional `r nrow(chargeEnds) - nrow(chargeBegins)` instances of the charge ending than there are of the charge beginning may be due to the first instance of data collection occurring during mid-charge for some vehicles. 

```{r calculateChargeDurations}
# select the observations which we've flagged as first & last in a sequence of charging
firstLastDT <- dt[chargeFlag == "first" | chargeFlag == "last"]
# flag the first of a pair
firstLastDT <- firstLastDT[, pairOK := ifelse(chargeFlag == "first" & shift(chargeFlag == "last", type = "lead"), "Pair start", NA)]
# flag the second of a pair
firstLastDT <- firstLastDT[, pairOK := ifelse(chargeFlag == "last" & shift(chargeFlag == "first"), "Pair end", pairOK)]
# calculate the time diff between all obs
firstLastDT <- firstLastDT[, pairDuration := difftime(time1 = dateTime, time2 = shift(dateTime), units = c("mins"))]
# we only want the time difference which was calculated for an obs where pairOK == "Pair end". This should also be where chargeFlag == "last" _except_ for where we have no 'first' (e.g. at start of data)
# note that we will still have pairs that bridge 00:00 which will give us -ve values
# if we have a -ve value then we need to change the calculation to add the time
# up to midnight from the start to the time after midnight to the end
firstLastDT <- firstLastDT[pairOK == "Pair start" & shift(pairDuration < 0, type = "lead"), 
                           toMidnight := difftime(time1 = as.hms("23:59:59"), time2 = time)]
firstLastDT <- firstLastDT[pairOK == "Pair end" & pairDuration < 0, 
                           afterMidnight := difftime(time1 = time, time2 = as.hms("00:00:00"), units = c("mins"))]
firstLastDT <- firstLastDT[, pairDurationFix := shift(toMidnight) + afterMidnight]
firstLastDT <- firstLastDT[, pairDurationFinal := ifelse(pairDuration <0,
                                                         pairDurationFix,
                                                         pairDuration)]

```

The charge duration was then calculated as being the time duration between each pair of "charge begins" and "charge ends" flags.

Figure \@ref(fig:durationHist) shows the overall distribution of all charging sequences. Clearly there are very small and a few very large values for Standard Charges, but this is not the case for Fast charges.

```{r durationHist, fig.cap="Duration of charging sequences"}

ggplot2::ggplot(firstLastDT[pairOK == "Pair end"], 
                aes(x = pairDurationFinal)) +
  geom_histogram(binwidth = 5) +
  facet_wrap(chargeType ~ ., scales = "free") +
  labs(x = "Minutes")
```

Table \@ref(tab:durationDescTable) shows the overall distributions and indicates the extent to which the means are skewed by the very small and a few very large values shown in Figure \@ref(fig:durationHist).

```{r durationDescTable}
t <- firstLastDT[pairOK == "Pair end", 
                 .(N = .N,
                   mean = mean(pairDurationFinal),
                   median = median(pairDurationFinal),
                   min = min(pairDurationFinal),
                   max = max(pairDurationFinal)), 
                 keyby = .(chargeType)]
kableExtra::kable(t, 
                  caption = "Duration of all charge sequences by charge type (minutes)", digits = 2) %>%
  kable_styling()
```

Figure \@ref(fig:shortDuration) shows the distribution of very short charging sequences. As we can see these appear to be generally less than 8 minutes in length for Standard Charges.

```{r shortDuration, fig.cap="Duration of charging sequences < 10 minutes"}

ggplot2::ggplot(firstLastDT[pairOK == "Pair end" & pairDurationFinal < 10], 
                aes(x = pairDurationFinal)) +
  geom_histogram(binwidth = 1) +
  facet_grid(chargeType ~ ., scales = "free") +
  labs(x = "Minutes")
```

Table \@ref(tab:durationDescTableReduced) shows the same descriptive statistics but for all sequences of greater than 8 minute duration. Now we can see that the mean and median durations for Standard Charge sequences are closer to one another.

```{r durationDescTableReduced}
t <- firstLastDT[pairOK == "Pair end" & pairDurationFinal > 8, 
                 .(N = .N,
                   mean = mean(pairDurationFinal),
                   median = median(pairDurationFinal),
                   min = min(pairDurationFinal),
                   max = max(pairDurationFinal)), 
                 keyby = .(chargeType)]
kableExtra::kable(t, 
                  caption = "Duration of charge sequences > 8 minutes by charge type (minutes, )", digits = 2) %>%
  kable_styling()
```


Manual inspection of the data showed that these short-duration charging "events" generally occurred near the end of a longer-duration charging event. It appeared that once the vehicle had reached its highest state of charge, charging would intermittantly stop and start again. This is likely due to the behaviour of the charger once the battery was almost full. As these can not be considered truly independent charging events, they have been removed from the data for the rest of the analysis.

In addition to the myriad "small" charging duration values, a small amount of unreasonably long charging durations (longer than 100 hours for standard charging or longer than 14 hours for fast charging) were calculated. As these exceeded the expected charge durations of the most high capacity vehicles currently available, they were assumed to be anomalies and were discarded.


```{r removeBigAndSmallDurations}
# Remove overly large values
firstLastDT <- firstLastDT[pairDurationFinal < 6000]

# ATTN BEN do we do the following here or disply the plots below that depend on the small values and then remove them for further analysis?

# Remove standard charges of duration less than 8 mins
firstLastDT <- firstLastDT[!(firstLastDT$pairDurationFinal < 8 & firstLastDT$chargeType == "Standard charging"),]

# Remove (lone) fast charge of duration greater than 10 hours
firstLastDT <- firstLastDT[!(firstLastDT$pairDurationFinal > 840 & firstLastDT$chargeType == "Fast charging"),]

# ATTN BEN rather than specify ...[pairOK == "Pair end", ] for all the following firstLastDT$pairDurationFinal plots, should we just remove all rows that are not "Pair end"?

```

Figure \@ref(fig:longDuration) shows the distribution of charging sequences with the excessively long or short events removed. As we can see these appear to be generally less than 3 hours in length for Standard Charges.

All further duration-related analysis is conducted with these unreasonably long or short events removed from the data.


```{r longDuration, fig.cap="Duration of charging sequences with unreasonably long or short values removed"}

ggplot2::ggplot(firstLastDT[pairOK == "Pair end"], 
                aes(x = pairDurationFinal)) +
  geom_histogram() +
  facet_wrap(chargeType ~ ., scales = "free") +
  labs(x = "Minutes")
```


# Key Findings:

```{r keyFindings}
stdMedian <- median(chargingDT[chargeType == "Standard charging"]$charge_power_kw, na.rm = TRUE)
stdMean <- mean(chargingDT[chargeType == "Standard charging"]$charge_power_kw, na.rm = TRUE)
  
fastMedian <- median(chargingDT[chargeType == "Fast charging"]$charge_power_kw, na.rm = TRUE)
fastMean <- mean(chargingDT[chargeType == "Fast charging"]$charge_power_kw, na.rm = TRUE)
```

 * _Power supplied_: The median power supplied during a standard charging was `r round(stdMedian,2)`  kW. The mean was slightly higher at `r round(stdMean,2)` kW. Fast charging observations had a median of `r round(fastMedian,2)` kW (mean = `r round(fastMean,2)`);
  * _Charging duration_: Charging durations tended to fall into one of two groups - longer 'overnight' charges with a median of XX hours and shorter events during the day both at standard and fast charge rates with a median duration of XX hours. 
  * _Time of Day_: charging events were more frequent at specific times of the day and day of the week with more evening and over-night charging during weekdays and more day-time charging at weekends. The power demand also varied according to time of day and day of the week.

# Observed demand

Figure \@ref(fig:obsPower) shows the distribution of observed charging kW demand by inferred charge type. This plot shows that fast charges are relatively rare in the dataset whilst standard charges are much more common, and are concentrated around 1.8kW, 3kW and 6kW. 

```{r obsPower, fig.cap="Observed power demand distribution by charge type where charging observed"}
p <- ggplot2::ggplot(chargingDT, aes(x = charge_power_kw, fill = chargeType, binwidth = 0.1)) +
  geom_histogram() +
  facet_wrap(. ~ chargeType, scales = "free")

# now draw the plot with any fancy extras we want
p + labs(y = "Density",
       x = "Power (kW)") +
  guides(fill = guide_legend(title = "Charge type:")) +
  scale_fill_manual(values=cbgPalette) + # use colour-blind friendly palette
  theme(legend.position = "bottom")
```

```{r stdChargeSizing}
stdQT <- quantile(chargingDT[chargeType == "Standard charging"]$charge_power_kw)
fastQT <- quantile(chargingDT[chargeType == "Fast charging"]$charge_power_kw)
```

75% of standard charging observations were `r round(stdQT[[2]],2)` kW or more but the figure was `r round(fastQT[[2]],2)` kW or more for fast charging

# Daily demand

```{r dailyPower, fig.cap="Observed power demand distribution by day of the week and charge type"}
p <- ggplot2::ggplot(filter(dt, chargeType == "Standard charging" | chargeType == "Fast charging"), aes(x = day_of_week, colour = chargeType, group = day_of_week)) +
  geom_boxplot(aes(y = charge_power_kw)) +
  facet_wrap(. ~ chargeType, scales= "free_y")

p + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  labs(x = "Day of week",
       y = "Power (kW)") +
  guides(colour = guide_legend(title = "Charge type:")) +
  scale_colour_manual(values=cbgPalette) + # use colour-blind friendly palette
  theme(legend.position = "bottom")
```

Figure \@ref(fig:dailyPower) shows the distribution of observed charging kW demand by day of the week. We can see that fast charging varies in demand more than standard charging does across days.


# Charging duration {#duration}



# Duration by time of day


```{r durationTimeBox, fig.cap="Duration by time of charging start"}

ggplot2::ggplot(firstLastDT[pairOK == "Pair end"], 
                aes(x = qHour, y = pairDurationFinal, group = qHour)) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) +
  geom_boxplot() +
  facet_grid(chargeType ~ ., scales = "free") +
  labs(x = "Time of Day",
       y = "Minutes")
```

```{r durationTimeMean, fig.cap="Mean duration (within quarter hours) by time of charging start for sequences > 8 minutes"}

plotDT <- firstLastDT[pairOK == "Pair end" & pairDurationFinal > 8,
                      .(meanDuration = mean(pairDurationFinal)),
                      keyby = .(qHour, chargeType)]
ggplot2::ggplot(plotDT, 
                aes(x = qHour, y = meanDuration, colour = chargeType)) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) +
  geom_point() +
  labs(x = "Time of Day",
       y = "Minutes")
```


```{r meanDurationTable}
kableExtra::kable(t, caption = "Mean duration of charge events by charge type")
```

```{r meanDurationPlot, fig.cap="Mean duration by time of charging start"}
plotDT <- firstLastDT[pairOK == "Pair end", .(meanDuration = mean(pairDurationFinal, na.rm = TRUE)), keyby = .(chargeType, dateTime)]
```


# Time of charging

```{r chargeTime, fig.cap="Count of observed charging events by type, day of week and time"}
plotDT <- chargingDT[, .(count = .N), keyby = .(qHour, chargeType, day_of_week)]

# make a weekend facet label
plotDT <- plotDT[, weekEnd := "Weekend"]
plotDT <- plotDT[day_of_week != "Saturday" & day_of_week != "Sunday", weekEnd := "Week day"]

p <- ggplot2::ggplot(plotDT, aes(x = qHour, y = count, colour = day_of_week)) +
  geom_line() +
  facet_grid(weekEnd ~  chargeType, scales = "free_y")
  
p + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  labs(x = "Time of day",
       y = "Count") +
  guides(colour = guide_legend(title = "Day of week:")) +
  scale_colour_manual(values=cbgPalette) + # use colour-blind friendly palette
  theme(legend.position = "bottom")
```

Figure \@ref(fig:chargeTime) shows the distribution of observed charging by time of day and day of the week. Aggregating counts in this way emphasises the times at which charging most commonly occurs and we can see...
 
Fig: profile of median charging demand by time of day and day of the week  faceted by at home vs not at home

Charging demand varies somewhat by time of day and day of the week. Weekdays show ... whilst weekends show. Saturdays and Sundays vary with... 

```{r boxplotCharging, fig.cap="Boxplot of charging timing by charge rate"}
p <- ggplot2::ggplot(subset(df, chargeType %in% "Standard charging"), 
                     aes(x = qHour, group = qHour, y = charge_power_kw)) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) +
  scale_colour_manual(values=cbbPalette) + # use colour-blind friendly palette
  geom_boxplot() # <- make the plot in an object first

p + labs(x = "Time of Day", y = "Power (kW)")
```









```{r plot3, fig.cap="Boxplot of charging timing by charge rate"}
p <- ggplot2::ggplot(subset(dt, chargeType %in% "Fast charging"), 
                     aes(x = qHour, group = qHour, y = charge_power_kw)) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) +
  scale_colour_manual(values=cbbPalette) + # use colour-blind friendly palette
  geom_boxplot() # <- make the plot in an object first

p + labs(x = "Time of Day", y = "Power (kW)")
```

```{r plot2, fig.cap="Boxplot of charging timing"}
p <- ggplot2::ggplot(dt, aes(x = qHour, group = qHour, y = charge_power_kw)) +
  guides(colour = guide_legend(title = "Vehicle:")) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) +
  scale_colour_manual(values=cbbPalette) + 
  geom_boxplot() 

p + labs(x = "Time of Day", y = "Power (kW)")
```



```{r ggjoyplotTimeChargingBegins, fig.cap="Time charging begins"}
ggplot(chargeBegins,aes(x=qHour, y= forcats::fct_rev(day_of_week))) +
  geom_density_ridges(rel_min_height = 0.01) +        # removes tails
  scale_x_discrete(expand = c(0.01, 0)) +  # removes cutoff top
  labs(x = "Hour", y = "Day") 
# Not sure how to get time on x axis (or if I want to)
# ATTN BEN I don't think this joy plot gives any more information than we get from the following 2 density plots. Delete this entire block if you agree.
```

```{r chargeBeginsWeekday, fig.cap="Density plot of charging start times during weekdays"}
p <- ggplot(chargeBegins[chargeBegins$weekday == "Weekday", ], aes(x = qHour, fill = chargeType)) +
  geom_density(alpha = 0.3) 
  facet_grid(~weekday)
p + labs(x = "Time", fill = "Charge type")

```

```{r chargeBeginsWeekend, fig.cap="Density plot of charging start times during weekends"}
p <- ggplot(chargeBegins[chargeBegins$weekday == "Weekend", ], aes(x = qHour, fill = chargeType)) +
  geom_density(alpha = 0.3) 
p + labs(x = "Time", fill = "Charge type")

```

```{r chargeEndsWeekday, fig.cap="Density plot of charging end times during weekdays"}
p <- ggplot(chargeEnds[chargeEnds$weekday == "Weekday", ], aes(x = qHour, fill = chargeType)) +
  geom_density(alpha = 0.3) 
p + labs(x = "Time", fill = "Charge type" )

```


```{r chargeEndsWeekend, fig.cap="Density plot of charging end times during weekends"}
p <- ggplot(chargeEnds[chargeEnds$weekday == "Weekend", ], aes(x = qHour, fill = chargeType))  +
  scale_colour_manual(values=cbbPalette)+
  geom_density(alpha = 0.3) +
  facet_grid(~weekday)
p + labs(x = "Time", fill = "Charge type")

```



At home charging events tended to begin at HH:MM during weekdays and HH:MM at weekends. 
_We can get "Slow" charging events rather than "home"_

Standard charging has a noticeably different profile to charging patterns for fast charges. It suggests that it is common for plug-in vehicle owners to charge overnight at home, and perhaps use the more powerful public chargepoints to top up during the day.

> Discuss any other patterns

# State of charge

The duration of charging events (see Section \@ref(duration)) suggests that EVs may be 'plugged in' at home (and elsewhere) for considerable durations. 


```{r SoCplot1, fig.cap= "Value of state of charge at beginning of charge"}
p <- ggplot(data=chargeBegins, aes(chargeBegins$SoC_percent)) + geom_histogram(bins = 20)
p + labs(x = "State of charge when charging begins (%)")
ggsave("~/EVBB/plots/SOC_when_charging_begins.png")
```
As can be seen in Figure \@ref(fig:SoCplot1), using the originally defined "charge begins" data we have the majority of charges beginning while the state of charge is above 90%. This is likely due to the manner in which the charger regularly turns off and on again near the end of the charging cycle.

Figure \@ref(fig:SoCplot2) shows the state of charge values when charge begins but with state of charge greater than 90% removed from the data. The figure shows that many vehicles arrive home with greater than 50% charge remaining. This indicates that charging may be delayed until early the following morning (during low aggregate electricity demand) while providing enough "back up" state of charge to allow for small evening trips if necessary. Alternatively, the battery may be able to transfer energy to the home during the evening grid peak as a form of demand response. 

```{r SoCplot2, fig.cap= "Value of state of charge at beginning of charge (>90% values removed)"}
tmp <- chargeBegins[chargeBegins$SoC_percent < 90, ]
p <- ggplot(data=tmp, aes(tmp$SoC)) + geom_histogram(bins = 20)
p + labs(x = "State of charge when charging begins (%)")
ggsave("~/EVBB/plots/SOC_when_charging_begins.png")
tmp <- NULL
```





Fig: Mean state of battery charge at the first 'at home' charging observation by hour and day of the week _No "at home" data with SOC_

> should show the timing of 'coming home' battery state?

Fig: Distribution of duration of charge events starting 'at home' in the evening (by day of the week) _Duration difficult to accurately determine without date due to charging occurring through the night_

The figure shows that vehicles may then be available for further demand response and/or re-charging for up to XX hours from this point.

> Discuss any other patterns

