---
title: "Analysis of electric vehicle usage patterns in New Zealand"
subtitle: "Statistical Report"
author: "Rafferty Parker and Ben Anderson (University of Otago)"
date: 'Last run at: `r Sys.time()`'
output:
  bookdown::pdf_document2:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
  bookdown::html_document2:
    code_folding: hide
    fig_caption: yes
    number_sections: yes
    self_contained: no
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    fig_caption: yes
    toc: yes
    toc_depth: 2
always_allow_html: yes
#bibliography: EVBBmendeleyrefs.bib  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(readr) # tidy data reading
library(data.table) # cos we like data.table (you may not in which case dplyr is fine :-)
library(lubridate) # for data & time manip
library(hms) # for hh:mm:ss if we need it
library(ggplot2) # fancy plots
library(ggjoy)
library(dplyr) # for filter
library(forcats) # used to reverse days of week in joy plots
library(knitr) # for knitting
library(kableExtra) # for extra kable

# colour blind palettes for charts
# http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#a-colorblind-friendly-palette
# with grey
cbgPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
# with black
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# data file to use
file <- "EVBB_processed_all_v1.0_20180125.csv"
# for Mac
user <- Sys.info()[[7]]
if(user == "ben"){
  dPath <- "/Volumes/hum-csafe/Research Projects/GREEN Grid/externalData/flipTheFleet/safe/testData/2019_01_25/"
  dFile <- paste0(dPath, file, ".zip") # use zipped data
  if(!file.exists(dFile)) {
    # we probably don't have the HCS mounted so switch to local
    dPath <- "~/Data/NZ_GREENGrid/ftf/"
    dFile <- paste0(dPath, file, ".zip")
  }
} else {
  # for Xubuntu:
  dPath <- "/run/user/1001/gvfs/smb-share:server=storage.hcs-p01.otago.ac.nz,share=hum-csafe,user=student%5Cparra358/Research Projects/GREEN Grid/externalData/flipTheFleet/safe/testData/2019_01_25/"
  dFile <- paste0(dPath, "EVBB_processed_all_v1.0_20180125.csv")
}

print(paste0("Using ", dFile))
rawDF <- readr::read_csv(dFile) # creates a tidyverse tibble https://www.tidyverse.org/articles/2018/01/tibble-1-4-1/

```


```{r dataPrep1}
df <- rawDF # so can always re-create df without having to re-load data
# don't do anything else here to avoid confusion
```

```{r dataPrep2}
#Combine date and time columns into POSIXct datetime
df$dateTime <- lubridate::as_datetime(paste0(df$date, df$time))

# set correct order for days of the week
df$day_of_week <- ordered(df$day_of_week, levels=c("Monday", "Tuesday", "Wednesday",
                                                   "Thursday", "Friday", "Saturday", "Sunday"))
# set charge type
df$chargeType <- ifelse(df$charge_power_kw > 0, "Standard charge", NA) 
df$chargeType <- ifelse(df$charge_power_kw >= 7, "Fast charge", df$chargeType)
df$chargeType <- ifelse(is.na(df$chargeType), "Not charging", df$chargeType) # not charging

# set charge type order so charts make sense from left (std) to right (fast)
df$chargeType <- cut(df$charge_power_kw, c(-Inf, 0.01, 7, Inf), labels = c('Not charging', 'Standard charging', 'Fast charging'))
df$chargeType <- factor(df$chargeType, ordered = TRUE)

# Rename vehicle ids to something more user-friendly
df$dvID <- factor(df$id, ordered = TRUE)
levSeq <- seq(1:length(levels(df$dvID)))
levSeqChar <- as.character(levSeq)
df$dvID <- factor(df$dvID,
  labels = levSeqChar)
df$dvID <- as.character(df$dvID)
df$dvID <- paste("Vehicle", df$dvID, sep = " ")

names(df)[names(df) == 'state_of_charge_percent'] <- 'SoC_percent'

df$qHour <- hms::trunc_hms(df$time, 15*60) # truncate to previous 15 min
#df$qHour <- format(as.POSIXct(hms::trunc_hms(df$time, 15*60)), "%H:%M")

# Month as ordered factor
df$month <- factor(df$month, ordered = TRUE, levels = c("Jan", "Feb", "Mar", "Apr", "May",
                                                        "Jun", "Jul", "Aug", "Sep", "Oct",
                                                        "Nov", "Dec"))


# Create factor for weekdays/weekends
weekdays1 <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
df$weekday <- factor((df$day_of_week %in% weekdays1), 
                   levels = c(TRUE, FALSE), labels = c('Weekday', 'Weekend'), ordered = TRUE)

# removal of silly state of charge percentage values
df$SoC_percent[df$SoC_percent > 100] <- NA
df$SoC_percent[df$SoC_percent < 0] <- NA

# removal of silly charge_power_kw values
# "...charging stations are being developed with capacities of 120kW in New Zealand"
# (Concept Consulting report)
df$charge_power_kw[df$charge_power_kw > 120] <- NA
```

```{r dataTable stuff, results="hide"}
dt <- as.data.table(df) # creates a data.table for fast data crunching

# Remove vehicles with all-zero charging values
summaryDT <- dt[, .(mean = mean(charge_power_kw), sd = sd(charge_power_kw), nObs = .N), keyby = .(dvID)]
includeDT <- summaryDT[mean != 0]
setkey(includeDT, dvID)
setkey(dt, dvID)
finalDT <- dt[includeDT]
dt <- finalDT

# Create charge flag
dt <- dt[, chargeFlag := ifelse(shift(charge_power_kw) == 0 & charge_power_kw > 0 & shift(charge_power_kw, type = "lead") > 0,
                                "first", "Not charging"), by = id] 
# the previous value was 0 but this value and the next value > 0

dt <- dt[, chargeFlag := ifelse(shift(charge_power_kw) > 0 & charge_power_kw > 0, 
                                "charging", chargeFlag), by = id] 
# previous value > 0, this value > 0

dt <- dt[, chargeFlag := ifelse(shift(charge_power_kw == 0, type = "lead") & charge_power_kw > 0 & shift(charge_power_kw) > 0, 
                                "last", chargeFlag), by = id] 
# next value = 0, this value and the previous value > 0



dt$chargeFlag <- ordered(dt$chargeFlag, levels=c("first", "charging", "last"))
 table(dt$chargeFlag, useNA = "always")

# ATTN BEN - delete following lines if not necessary
#dt <- dt[ , `:=`( chargeCount = .N ) , by = chargeFlag ]

#dt <- dt[, obsDiffTime := difftime(dateTime,shift(dateTime)), by = id] # time since previous observation (within id)

#dt <- dt[, obsDiffSecs := as.numeric(obsDiffTime)] # seconds since previous observation (within id) - could include reset to 0 after midnight
  
chargingDT <- dt[charge_power_kw > 0] # select just charging
```

# Key Findings:{#keyFindings}
Based on a relatively small sample of 44 domestic electric vehicles provided by our research partner and monitored over 8 months from April 2018 to January 2019. The recorder provided measurements at 1 minute frequency of charging power and battery charge state.

 * _Power supplied_: The median power supplied during a standard charging event was 1.78 kW. The mean was slightly higher at 2.12 kW. Fast charging observations had a median of 30.84 kW (mean = 30.68kW);
  * _Charging duration_:Charging durations tended to fall into one of two groups. Longer ‘standard’ charges had a median duration of 0.06 hours and a mean duration of 1.69 hours. High power “fast” charge events had a median duration of 12.47 minutes and a mean duration of 13.87 minutes.
  * _Time of day_: Standard charging events tended to begin around 10pm, suggesting the drivers in our dataset utilise timers to take advantage of off-peak electricity. Fast charging events tended to begin at 11:30am on weekdays and 1pm during weekends.
  * _State of charge_: Many drivers begin recharging with greater than 50% charge still remaining in the battery.
  
In concurrence with the 2018 Concept Consulting report[@ConceptConsulting2018], these findings suggest that any negative effects electric vehicles may have on the evening national electricity grid peaks should be mitigable through "smart" charging methods. In addition, our analysis indicates that this is already occurring to some extent. 

# Data information{#data}

## Background 

The data used has been provided by "Flip the Fleet", a community organisation that hopes to increase uptake of electric vehicles in New Zealand. Flip the Fleet have been collecting data on electric vehicle usage patterns, collected using Exact IOT Limited's [blackbox recorder](https://flipthefleet.org/ev-black-box/), a small electronic  device that connects to the vehicle's internal computer and sends detailed data about the battery health, consumption, speed, etc.

The data used was collected from `r  length(levels(as.factor(dt$dvID)))` vehicles, in total consisting of `r nrow(dt)` data points, over 8 months (April 2018 - January 2019). The recorder provided measurements at 1 minute frequency of charging power and battery charge state. It is important to note that these vehicles are driven by "early adopters" who have opted to collect their vehicle usage data, and therefore the data may not be a precise representation of the usage patterns of current or future EV drivers.

Due to privacy considerations, the data is not publicly available.

```{r establishment of chargeBegins and chargeEnds dataframes}
chargeBegins <- chargingDT[chargingDT$chargeFlag == "first" , ]
chargeEnds <- chargingDT[chargingDT$chargeFlag == "last", ]
```


```{r calculateChargeDurations}
# select the observations which we've flagged as first & last in a sequence of charging
firstLastDT <- dt[chargeFlag == "first" | chargeFlag == "last"]
# flag the first of a pair
firstLastDT <- firstLastDT[, pairOK := ifelse(chargeFlag == "first" & shift(chargeFlag == "last", type = "lead"), "Pair start", NA)]
# flag the second of a pair
firstLastDT <- firstLastDT[, pairOK := ifelse(chargeFlag == "last" & shift(chargeFlag == "first"), "Pair end", pairOK)]
# calculate the time diff between all obs
firstLastDT <- firstLastDT[, pairDuration := difftime(time1 = dateTime, time2 = shift(dateTime), units = c("mins"))]
# we only want the time difference which was calculated for an obs where pairOK == "Pair end". This should also be where chargeFlag == "last" _except_ for where we have no 'first' (e.g. at start of data)
# note that we will still have pairs that bridge 00:00 which will give us -ve values
# if we have a -ve value then we need to change the calculation to add the time
# up to midnight from the start to the time after midnight to the end
firstLastDT <- firstLastDT[pairOK == "Pair start" & shift(pairDuration < 0, type = "lead"), 
                           toMidnight := difftime(time1 = as.hms("23:59:59"), time2 = time)]
firstLastDT <- firstLastDT[pairOK == "Pair end" & pairDuration < 0, 
                           afterMidnight := difftime(time1 = time, time2 = as.hms("00:00:00"), units = c("mins"))]
firstLastDT <- firstLastDT[, pairDurationFix := shift(toMidnight) + afterMidnight]
firstLastDT <- firstLastDT[, pairDurationFinal := ifelse(pairDuration <0,
                                                         pairDurationFix,
                                                         pairDuration)]

```


```{r removeBigAndSmallDurations}
# Remove overly large values
firstLastDTfinal <- firstLastDT[pairDurationFinal < 6000]

# ATTN BEN do we do the following here or disply the plots below that depend on the small values and then remove them for further analysis?

# Remove standard charges of duration less than 8 mins
firstLastDTfinal <- firstLastDTfinal[!(firstLastDTfinal$pairDurationFinal < 8 & firstLastDTfinal$chargeType == "Standard charging"),]

# Remove (lone) fast charge of duration greater than 10 hours
firstLastDTfinal <- firstLastDTfinal[!(firstLastDTfinal$pairDurationFinal > 840 & firstLastDTfinal$chargeType == "Fast charging"),]

# ATTN BEN rather than specify ...[pairOK == "Pair end", ] for all the following firstLastDT$pairDurationFinal plots, should we just remove all rows that are not "Pair end"?

```

Figure \@ref(fig:longDuration) shows the charging durations.  

```{r longDuration, fig.cap="Duration of charging sequences with unreasonably long or short values removed"}

ggplot2::ggplot(firstLastDTfinal[pairOK == "Pair end"], 
                aes(x = pairDurationFinal)) +
  geom_histogram(binwidth = 5) +
  facet_wrap(chargeType ~ ., scales = "free") +
  labs(x = "Minutes")
ggsave("plots/charge_duration_histogram_extrema_removed.png")
```



```{r keyFindings}
stdMedian <- median(chargingDT[chargeType == "Standard charging"]$charge_power_kw, na.rm = TRUE)
stdMean <- mean(chargingDT[chargeType == "Standard charging"]$charge_power_kw, na.rm = TRUE)
  
fastMedian <- median(chargingDT[chargeType == "Fast charging"]$charge_power_kw, na.rm = TRUE)
fastMean <- mean(chargingDT[chargeType == "Fast charging"]$charge_power_kw, na.rm = TRUE)

fastDurMedian <- median(firstLastDTfinal[pairOK == "Pair end" & chargeType == "Fast charging"]$pairDurationFinal, 
                        na.rm = TRUE)
standardDurMedian <- median(firstLastDTfinal[pairOK == "Pair end" & chargeType == "Standard charging"]$pairDurationFinal, 
                        na.rm = TRUE)

fastDurMean <- mean(firstLastDTfinal[pairOK == "Pair end" & chargeType == "Fast charging"]$pairDurationFinal, 
                        na.rm = TRUE)
standardDurMean <- mean(firstLastDTfinal[pairOK == "Pair end" & chargeType == "Standard charging"]$pairDurationFinal, 
                        na.rm = TRUE)
```


# Observed demand

Figure \@ref(fig:obsPower) shows the distribution of observed charging kW demand by inferred charge type. This plot shows that fast charges are relatively rare in the dataset whilst standard charges are much more common, and are mostly concentrated around 1.8kW and 3kW, with a smaller concentration around 6kW.

```{r obsPower, fig.cap="Observed power demand distribution by charge type where charging observed"}
p <- ggplot2::ggplot(chargingDT, aes(x = charge_power_kw, fill = chargeType, binwidth = 0.1)) +
  geom_histogram() +
  facet_wrap(. ~ chargeType, scales = "free")
p + labs(y = "Density",
       x = "Power (kW)") +
  guides(fill = guide_legend(title = "Charge type:")) +
  scale_fill_manual(values=cbgPalette) + # use colour-blind friendly palette
  theme(legend.position = "bottom")
```

```{r stdChargeSizing}
stdQT <- quantile(chargingDT[chargeType == "Standard charging"]$charge_power_kw)
fastQT <- quantile(chargingDT[chargeType == "Fast charging"]$charge_power_kw)
```

75% of standard charging observations were `r round(stdQT[[2]],2)` kW or more but the figure was `r round(fastQT[[2]],2)` kW or more for fast charging

# Charging duration {#duration}

Figure \@ref(fig:durationTimeMean) show that the duration of standard charging events by event end time drops significantly for events ending around 9:45am. This may indicate that people are plugging in after returning home from a school run or other morning activity, even though the battery is still close to full capacity. It may also suggest that those who plug in shortly after 9:45am but do not have a high battery state of charge are only "topping up", and take the vehicle out again before charging is fully complete. Duration of fast charge events by event end time appear to be more randomly distributed, although very few events were recorded between midnight and 7am. This, along with the comparatively low number of recorded fast charge events indicated in Fig. \@ref(fig:obsPower) suggests that drivers utilize fast charging only "as necessary" to ensure they have enough battery capacity to complete their journey.


```{r durationTimeMean, fig.cap="Mean duration (within quarter hours) by time of charging end"}

plotDT <- firstLastDTfinal[pairOK == "Pair end" & pairDurationFinal > 8,
                      .(meanDuration = mean(pairDurationFinal)),
                      keyby = .(qHour, chargeType)]
ggplot2::ggplot(plotDT, 
                aes(x = qHour, y = meanDuration, colour = chargeType)) +
  theme(legend.position = "bottom", axis.text.x = element_text()) +
  geom_point() +
  labs(x = "Time of Day",
       y = "Minutes", col = "")
```


```{r meanDurationTable}
kableExtra::kable(t, caption = "Mean duration of charge events by charge type")
```

```{r meanDurationPlot, fig.cap="Mean duration by time of charging start"}
plotDT <- firstLastDTfinal[pairOK == "Pair end", .(meanDuration = mean(pairDurationFinal, na.rm = TRUE)), keyby = .(chargeType, dateTime)]
```

# State of charge{#SoC}

The state of charge is the percentage of energy still available to be used in the battery. In future, electric vehicles may be able to discharge any remaining battery charge as electricity into the grid, a process known as vehicle to grid (V2G) energy transfer. This may allow electric vehicles to have a net beneficial effect on the grid, reducing the evening peaks by providing electricity to the home during this period, and then recharging later in the evening or early the next morning when peak demand has diminished.

This section provides an indication of the state of charge of electric vehicles upon charging, so that the potential of V2G technology can be assessed.



Figure \@ref(fig:SoCplot2) shows the state of charge values when charge begins. The figure indicates that many vehicles begin charging despite having greater than 50% charge remaining.

```{r SoCplot2, fig.cap= "Value of state of charge at beginning of charge (>90% values removed)"}
tmp <- chargeBegins[chargeBegins$SoC_percent < 90, ]
p <- ggplot(data=tmp, aes(tmp$SoC)) + geom_histogram(bins = 20)
p + labs(x = "State of charge when charging begins (%)")
ggsave("plots/SOC_when_charging_begins.png")
tmp <- NULL
```


# Time charging begins

Figures \@ref(fig:chargeBeginsWeekday and \@ref(fig:chargeBeginsWeekday) show a distribution plot of charging start times.


```{r chargeBeginsWeekday, fig.cap="Density plot of charging start times during weekdays"}
p <- ggplot(chargeBegins[weekday == "Weekday" & SoC_percent < 90, ], aes(x = qHour, fill = chargeType)) +
  geom_density(alpha = 0.3) +
  theme(legend.position = "bottom") 
  facet_grid(~weekday)
p + labs(x = "Time", fill = "Charge type")
ggsave("plots/time_charging_begins_weekday.png")

```


```{r chargeBeginsWeekend, fig.cap="Density plot of charging start times during weekends"}
p <- ggplot(chargeBegins[weekday == "Weekend" & SoC_percent < 90, ], aes(x = qHour, fill = chargeType)) +
  theme(legend.position = "bottom") +
  geom_density(alpha = 0.3) 
p + labs(x = "Time", fill = "Charge type")
ggsave("plots/time_charging_begins_weekend.png")

```

Standard charging has a noticeably different profile to charging patterns for fast charges. It suggests that it is common for plug-in vehicle owners to charge overnight at home, and perhaps use the more powerful public charge points to top up during the day.

Standard charging events most commonly began around 10pm during both weekdays and weekends. As it seems unlikely that this is due to vehicle drivers returning home at this hour, this effect may be due to drivers setting the charger on a timer to take advantage of cheaper "off-peak" electricity times, which frequently begin around 10pm.

Fast charging events tended to begin at 11:30am on weekdays and 1pm during weekends.  


```{r chargeTime, fig.cap="Count of observed charging events by type, day of week and time"}

plotDT <- chargingDT[, .(count = .N), keyby = .(qHour, day_of_week)]

# make a weekend facet label
plotDT <- plotDT[, weekEnd := "Weekend"]
plotDT <- plotDT[day_of_week != "Saturday" & day_of_week != "Sunday", weekEnd := "Week day"]

p <- ggplot2::ggplot(plotDT, aes(x = qHour, y = count, colour = day_of_week)) +
  geom_line() +
  facet_grid(~weekEnd, scales = "free_y")
  
p + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  labs(x = "Time of day",
       y = "Count") +
  guides(colour = guide_legend(title = "Day of week:")) +
  scale_colour_manual(values=cbgPalette) + # use colour-blind friendly palette
  theme(legend.position = "bottom")
```

Figure \@ref(fig:chargeTime) shows the distribution of observed charging by time of day and day of the week. This figure indicates greatest charging occurance between the hours of 8pm and 8am, with very low occurrences of charging during morning and evening grid peaks.

# Summary

In the data provided for this study, most charging occurs at home using either a 1.8kw or 3kW charger, and commonly occurs through the night as opposed to during current grid peaks. In addition, many vehicles begin charging with significant battery capacity remaining, providing them with the ability to provide vehicle to grid energy transfer should that technology become widely available. 

If later adopters of electric vehicles can be induced to follow the same "smart" charging patterns as those displayed in our data sample, we propose the effects that electric vehicles have on the electricity grid may not be particularly negative.

# References